Maybe Slipstreams

### 2014-03-16 paper.tex
DSL Stuffff

\documentclass[twocolumn,11pt,english]{article}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{cite}
\usepackage[T1]{fontenc}
\usepackage[margin=1.0in]{geometry}
\lstset{language=Haskell}

@@

\section{Introduction}
\subsection{What is FRP}
Functional Reactive Programming, introduced by Fran \cite{ElliottHudak97:Fran}, is a programming paradigm based on time varying values. The fundamental idea behind it is to allow the programmer to express than ``what'' of a program, and the ``how'' of the program be infered. FRP operates by a series of \textit{behaviors}, which are time varying values that react with one another to form a program.

@@

\subsection{Problems with FRP}
We solve the problem of interacting with time-varying values can be solved using stream-based FRP, where each computation sends a stream instead of a signal value. A stream is essentially a complete history of the values of a particular computation. Any other computation that needs the current value, or a value from the past, can look at the stream and find the appropriate values. This leads to additional problems like memory usage growing out of control as new values are continually added to the stream and not removed. In some systems garbage collection finds and removes values from streams that are no longer dependencies for other computations. In Elerea \cite{HighOrderStreams}, this is solved by only allowing access to the most recent values in streams.

@@

\section{Terminology}
\begin{description}
\item[element] Any value or computation that relies upon previous computations or values. An element is analogous to a node in a graph. The element is connected to various other elements by means of some function. All elements connected to element $e$ can be split into two sets $s_1$ (all elements supplying input to $e$) and $s_2$ (all elements recieving their input from $e$). $s_1$ is called the predecessor set of $e$ or $s_1 = pre(e)$, and $s_2$ is the successor set of $e$ or $s_2 = suc(e)$

@@

\item[layer] A layer is a set of elements in a network where the intersection of their mutual union of predecessor and successor sets is the empty set. More formally $\bigcup_{e_i \in L}{pre(e_i)} \cap \bigcup_{e_j \in L}{suc(e_j)} = \emptyset$ where $L$ is a layer. Breaking a network into layers is helpful because it helps identify dependencies in the network. It also helps introduce structure into the environment, which can help organize a network which may have been difficult to think about. It is not always possible to adhere to the definition of a layer, or is not practical to do so (even if it could) be done. For example, a feedback loop would be, by definition, a loop on the network, some way of passing a result back to a predecessor. This technique is useful for maintaining state in a network amongst other uses.

@@

\section{Our contributions}
Our contributions are basically jsut reworking the bind in a stream monad to map between layer sets. Also each value is wrapped in a maybe, revolutionary. This allows safe access to resources that have not yet been created in a dynamic network setting. It also allows for safe handling of volatile resources such as HTTP requests or accessing other error prone procedures. Convenience elements have been added such as the aggregator, which allows for dynamic seeming networks without any modification to the underlying structure. This is possible due to the addition of the maybe value to every stream.
\subsection{DSL Grammar}
\footnotesize
\begin{align*}
  Statement ::=& ~V \leftarrow Expression &\text{Variable assignment}\\
  |& ~Expression &\text{Single expression}\\
  Expression ::=& ~SignalNet &\text{A signal network}\\
  |& SimpleExpr &\text{An expression}\\
  SignalNet ::=& ~( Exprs ) \rightarrow SignalNet &\text{Compose layers}\\
   |& ~(Exprs ) &\text{Single layer} \\
  Exprs ::=& ~SimpleExpr ~,~ Exprs &\text{List of expressions}\\
  |& ~SimpleExpr &\text{Single expression}\\
  SimpleExpr ::=& ~L[``Javascript code''] &\text{Lift}\\
  |& ~ @V &\text{Get}\\
  |& ~ *V(SimpleExpr) &\text{Put}\\ 
  |& ~ V( Exprs ) &\text{Function call}\\
  V ::=& ~Word~Characters &\text{Variable}
\end{align*}
\normalsize
As you can see our grammar is exteremly simple and contributes basically 1 monad with bind. Thats it. PhD Please.

\subsection{Maybe Streams}
Merging streams and wormholes and making an implementation of the whole thing practical. Making it based in theory so its not a giant mess like bacon.js. Monad streams woo!

\subsection{Higher-order streams}
Streams composed with streams oh my. Gotta look out for those undefined values. The value of composed streams is the diagonal of the matrix formed by the stream values. Boop:

\begin{align*}
  \langle &\langle \mathbf{v_{00}} ~ v_{01} ~ v_{02} ~ v_{03} ~ ... ~ \rangle \\
  &\langle \perp ~ \mathbf{v_{11}} ~ v_{12} ~ v_{13} ~... ~ \rangle \\
  &\langle \perp ~ \perp ~ \mathbf{v_{22}} ~ v_{23} ~ ...  ~ \rangle \rangle
\end{align*}

@@

\section{Related work}

\subsection{Higher-order streams}
Work has already been done on higher-order streams \cite{HighOrderStreams}. Higher-order streams are nessisary for having a dynamic network. Dynamic networks are required

\subsection{Wormholes}
\cite{WinogradCort2012HS}
\section{Future work}
Probably some since we want to keep working on things. I bet it will have to do with extending the language definition and possibly defining streams better.

>>>>

\documentclass[twocolumn,11pt,english]{article}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{cite}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{color}
\usepackage[margin=1.0in]{geometry}
\lstset{language=Haskell}

@@

\section{Introduction}

\subsection{What is FRP}
Functional Reactive Programming, introduced by Fran \cite{ElliottHudak97:Fran}, is a programming paradigm based on time varying values. The fundamental idea behind it is to allow the programmer to express than ``what'' of a program, and the ``how'' of the program be infered. FRP operates by a series of \textit{behaviors}, which are time varying values that react with one another to form a program.

@@

\subsection{Problems with FRP}
We solve the problem of interacting with time-varying values using stream-based FRP, where each computation sends a stream instead of a signal value. A stream is essentially a complete history of the values of a particular computation. Any other computation that needs the current value, or a value from the past, can look at the stream and find the appropriate values. This leads to additional problems like memory usage growing out of control as new values are continually added to the stream and not removed. In some systems garbage collection finds and removes values from streams that are no longer dependencies for other computations. In Elerea \cite{HighOrderStreams}, this is solved by only allowing access to the most recent values in streams.

@@

\section{FRP Terminology}

\begin{description}

@@

@@

\section{Network Dynamism through Higher Order Streams}
In order to express the dynamic behavior that we need to implement GUIs, the underlying signal network needs to be able to change itself over time. We choose to model this dynamism with higher order streams like Patai\cite{HighOrderStreams}. This method avoids the time and space leaks that were present in other FRP systems with first class signals/behaviors/events \cite{ElliottHudak97:Fran}.

Patai defines his model in terms of \textit{Denotational Design} developed by Conal Elliot \cite{elliott2009denotational}. The core idea Patai leverages is the \textit{Type Class Morphism}, or defining an interface by a combination of its \textit{type} and its \textit{properties}. By focusing on the interface, this design strategy allows implementations to be checked for correctness but to otherwise have freedom in their methodology.

The core type in Patai's stream model is \textit{Stream a}, representing a stream of values over time. This is analagous to a Behavior, and in fact has the same type signature: 
\begin{equation}
  \textbf{type}~\textit{Stream a} = \mathbb{N} \rightarrow a
\end{equation}
Any  s::\textit{Stream a} can be thought of as the stream of values:
\begin{equation}
  \langle s_0 s_1 s_2 ... s_n \rangle ~where~ s_i :: a
\end{equation}
Initially Patai models these Streams using a \textit{cons} constructor that represents a delay of one timestep. However, if we use this stream definition and consider a higher order stream of type \textit{Stream (Stream a)}, we end up with quadratic performance for \textit{join} when we attempt to extract a value from the \textit{Stream (Stream a)}. This is fairly obvious if we try to visualize the higher order stream, since the values of $join~s$ where $s$ :: \textit{Stream (Stream a)} are the main diagonal of the stream of streams:

\begin{align*}
    \langle &\langle \mathbf{v_{00}} ~ v_{01} ~ v_{02} ~ ... ~ \rangle \\
    &\langle v_{10} ~ \mathbf{v_{11}} ~ v_{12} ~ ... ~ \rangle \\
    &\langle v_{20} ~ v_{21} ~ \mathbf{v_{22}} ~  ...  ~ \rangle \rangle
\end{align*}

To solve this problem, Patai moves away from the \textit{cons} definition of Streams. Instead, he creates streams and higher order streams from three primitives: $delay$, $start$, and $generator$. He also introduces the type $StreamGen$, which has the same signature as a $Stream$ but is only used to represent higher order streams.

\small
\begin{equation}
  \mathbf{type}~\mathit{StreamGen~a} = \mathbb{N} \rightarrow a
\end{equation}
\normalsize
The $delay$ function replaces cons in this new configuration. The difference is that instead of returning a $Stream$, it returns a $StreamGen$. It functions by returning its first argument at $t = 0$, and the first $Stream~a$ applied to the sample time otherwise. For intuition, we've also included the simplified type:
\small
\begin{align*}
dela&y :: a \rightarrow Stream~a \rightarrow StreamGen~(Stream~a)\\
dela&y :: a \rightarrow (\mathbb{N} \rightarrow a) \rightarrow (\mathbb{N} \rightarrow~(\mathbb{N} \rightarrow~a))\\
dela&y~x~s~t_{start}~t_{sample} =\\
&|~t_{start} = t_{sample} = x\\
&|~t_{start} < t_{sample} = s (t_{sample} - 1)\\
&|~otherwise ~~~~~~=\perp
\end{align*}
\normalsize

The $start$ function simply gives us a safe way to extract a stream from a stream generator. Given that $delay$ is undefined if $t_{start} > t_{sample}$, a universally safe time for $t_{start}$ is 0.
\small
\begin{align*}
&start :: StreamGen (Stream~a) \rightarrow Stream a\\
&start g = g~0
\end{align*}
\normalsize

However, given $t_{sample}$ there is another safe value: $t_{start} = t_{sample}$. This concept is embodied in the function $generator$: 
\small
\begin{align*}
&generator :: Stream~ (StreamGen a) \rightarrow Stream~a\\
&generator~g~t_{sample} = g~t_{sample}~t_{sample}
\end{align*}
\normalsize

TODO: DISCUSS THE FANCY EFFICIENCIES AND BIND AND JOIN ETC
\begin{align*}
  \langle &\langle \mathbf{v_{00}} ~ v_{01} ~ v_{02} ~ v_{03} ~ ... ~ \rangle \\
  &\langle \perp ~ \mathbf{v_{11}} ~ v_{12} ~ v_{13} ~... ~ \rangle \\
  &\langle \perp ~ \perp ~ \mathbf{v_{22}} ~ v_{23} ~ ...  ~ \rangle \rangle
\end{align*}

@@

\section{Real World Interfaces}
\subsection{Signals as Maybe Streams}
  Consider modeling a stream of data from a server to a browser GUI. There are two situations when the value of this stream might have an undefined value: If the connection between server and browser has not yet been initiated, or if it has been disrupted.

To model the two cases of \textit{disruption} and \textit{pre-existence}, we introduce what we call a $Signal$, which is roughly modeled as:
\begin{equation}
  \textbf{type}~\textit{Signal a} = Stream (Maybe~a)
\end{equation}

At any point in time, a $Signal$ may now be defined but have a $Nothing$ value. We can now be assured that any system behavior dependent on this value will act in a defined way even given possible disruption.

As in the given motivating example, we could now also reference the current value of a stream of data coming from the server before we ever initiate the connection.

\subsection{Wormholes}
\cite{WinogradCort2012HS}

\section{DSL}

A major roadblock in FRP adoption is its complexity. Understandably, FRP implementations attempt to be as expressive as possible. Our implementation, for example, extended the basic idea of Elerea and Elm with additional safety measures for real world interaction.

The goal of our DSL is to bridge the gap between a complex, fully expressive FRP system, and something that can be easily used to express a usable FRP program. It is intended to rest atop more complicated FRP libraries, enabling access for a new class of end users who want a quick and intuitive interface.

\subsection{DSL Components}

When implementing software with FRP, dependencies between values can be confusing. This is why in our DSL, we restrict dependencies by dividing Signal values into \textit{layers}. Layers are composed of \textit{elements}, representing time-varying values ($Signal$s) computed from previous values. To maintain the desired ease of understanding, elements can only depend on the values of elements from the previous layer.

\begin{figure}[h!]
  \includegraphics[scale=0.8]{layers.eps}
\caption{DSL Layers and Elements}
\label{fig:layers}
\end{figure}

In Figure~\ref{fig:layers}, arrows represent dependencies. Element $e_{22}$, for example, depends on $e_{11}$ and $e_{12}$. In this system, the output of the FRP program is defined as the tuple of the values of the elements of the last layer; The output of this FRP program would be $(e_{31}, e_{32}, e_{33})$.

\subsection{DSL Definitions}
\begin{description}
\item[element] Any time-varying value, or any computation that relies upon previous computations or values. An element is essentially a Signal Function that has arity from 0 to $n$ where $n$ is the size of the previous layer. All elements connected to element $e$ can be split into two sets: $s_1$ (all elements supplying input to $e$) and $s_2$ (all elements recieving their input from $e$). $s_1$ is called the predecessor set of $e$ or $s_1 = pred(e)$, and $s_2$ is the successor set of $e$ or $s_2 = succ(e)$

\item[layer] A layer is a set of elements in a network where the intersection of their mutual union of predecessor and successor sets is the empty set. More formally,
\begin{equation}
  \bigcup_{e_i \in L}{pred(e_i)} \cap \bigcup_{e_j \in L}{succ(e_j)} = \emptyset
\end{equation}
where $L$ is a layer. Breaking a network into layers is helpful because it helps identify dependencies in the network. It also helps introduce structure into the environment, which can help organize a network during its development. 
\end{description}
The DSL defined in this way cannot express many network topologies. For example, it cannot express feedback loops, and has no way of passing a result back to a predecessor. This technique is useful for maintaining state in a network, and there are many other practical uses for it.

\subsection{DSL Examples}
The goal of the DSL is to easily express \textit{networks} as we've defined them. To express the network from Figure~\ref{fig:layers}, we would write:
\begin{center}
$(e_{11}, e_{12}, e_{13})$
\\ $\rightarrow (e_{21}, e_{22})$
\\ $\rightarrow (e_{31}, e_{32}, e_{33})$
\end{center}
The elements in layer 1 represent functions of type $() \rightarrow Signal~a$, while the elements in all other layers represent a signal function of any arity, i.e.) $Signal~(a, b, ... n) \rightarrow Signal~z$

A program in the actual DSL would therefore have $source$s in the first layer and signal functions or sources elsewhere. Builtin functions of type $() \rightarrow Signal~a$ are prefixed by $@$ in our DSL for clarity.

In our DSL, $L[expr]$ represents the lifting of an anonymous function that takes as arguments the values of all elements in the previous layer, and whose body is $expr$. That is, 
\begin{center}
$L[\$1 * \$2] \equiv lift (\backslash \$1 ~\$2 \rightarrow \$1 * \$2)$
\end{center}
Using these constructs, a specific network with the same topology as Figure~\ref{fig:layers} might be:
\begin{center}
$(@mousePosX, @mousePosY, @time)$
\\ $\rightarrow (L[sum(\$1, \$2)], L[\$3 * 2])$
\\ $\rightarrow (L[\$1 / 2], L[\$1], L[\$3])$
\end{center}
\subsection{DSL Grammar}
\footnotesize
\begin{align*}
  Statement ::=& ~V \leftarrow Expression &\text{Variable assignment}\\
  |& ~Expression &\text{Single expression}\\
  Expression ::=& ~SignalNet &\text{A signal network}\\
  |& SimpleExpr &\text{An expression}\\
  SignalNet ::=& ~( Exprs ) \rightarrow SignalNet &\text{Compose layers}\\
   |& ~(Exprs ) &\text{Single layer} \\
  Exprs ::=& ~SimpleExpr ~,~ Exprs &\text{List of expressions}\\
  |& ~SimpleExpr &\text{Single expression}\\
  SimpleExpr ::=& ~L[``Javascript code''] &\text{Lift}\\
  |& ~ @V &\text{$SF :: () \rightarrow Signal~a$}\\
  |& ~ !V &\text{Get}\\
  |& ~ *V(SimpleExpr) &\text{Put}\\ 
  |& ~ V( Exprs ) &\text{Function call}\\
  V ::=& ~Word~Characters &\text{Variable}
\end{align*}
\normalsize

\subsection{Desugaring}
This DSL can be desugared to a base that is easily analyzed in the context of the FRP system we've constructed so far. Given the two layer network:
\small
\begin{align*}
(@time, @mousePosX) \rightarrow (L[\$1 * \$2], L[\$1 + \$2])
\end{align*}
\normalsize
If we introduce a function $mtup$ where 
\small
\begin{align*}
mtup~::~(Monad~m) \Rightarrow (m~a, ..., m~n) \rightarrow m (a, ..., n)
\end{align*}
\normalsize
We can desugar this specific example to:
\small
\begin{align*}
l1~=&~(time, mousePosX)
\\l2~=&
\\&(\backslash (\$1, \$2) \rightarrow return~ (\$1 * \$2, \$1 + \$2)) \gg = mtup~l1
\end{align*}
\normalsize
\section{Related work}

\section{Future work}
Probably some since we want to keep working on things. I bet it will have to do with extending the language definition and possibly defining streams better.
